---
title: "Chess Game Dataset"
author: "Austin Mangelson"
format: 
  html:
    code-fold: true
    code-tools: true
    embed-resources: true
editor: visual
execute: 
  warning: FALSE
  message: FALSE
---

```{r}
library(tidyverse)

chess <- readr::read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2024/2024-10-01/chess.csv')
```

Questions

-   most common opening move(s) for winners

-   average game length

-   **Rating Impact on Win Probability**:

    -   Analyze how the difference in ratings between players (white vs black) correlates with the probability of winning. You can also look at this for rated and unrated games separately.

<!-- -->

-   **Game Length vs Winner**:

    -   Compare the number of turns in games where the white player won vs the black player. This could reveal whether certain types of games (short or long) favor one color.

-   **Opening Strategy Effectiveness**:

    -   Evaluate the success rates of different chess openings (based on `opening_eco` or `opening_name`). You can check win rates and whether certain openings tend to lead to shorter or longer games.

-   **Time Control Influence**:

    -   Investigate how different time controls (given by `time_increment`) impact the number of turns and the win/loss ratio. Does faster time control favor higher-rated players more than slower ones?

-   **Opening Ply and Game Outcome**:

    -   Examine whether the number of opening moves (`opening_ply`) has any relationship with the outcome of the game. Do longer openings indicate more decisive games, or does it lead to draws?

-   **Moves Complexity**:

    -   Explore whether longer move sequences or games with more complex opening names lead to victories by higher-rated players. Use `moves` data to analyze depth and variability in the game.

-   **Victory Status Breakdown**:

    -   Analyze the breakdown of victory statuses (`victory_status`): checkmate, resignation, time, or stalemate, and correlate them with player ratings, game length, and openings.

-   **Rated vs Unrated Games**:

    -   Compare rated and unrated games for trends in victory status, number of turns, and moves to see if players behave differently in casual vs rated environments.

------------------------------------------------------------------------

## General Stats

```{r}
chess |> 
  summarise(count = n_distinct(game_id))
```

19,113 games

```{r}
chess |> 
  select(white_id, black_id) |> 
  pivot_longer(cols = everything(), names_to = "color", values_to = "player_id") |>
  distinct(player_id) |> 
  summarise(count = n())
```

15,635 players

```{r}
win_dist <- chess |> 
  group_by(winner) |> 
  summarise(count = n()) |> 
  arrange(desc(count))

win_dist$fraction <- win_dist$count / sum(win_dist$count)
win_dist$ymax <- cumsum(win_dist$fraction)
win_dist$ymin <- c(0, head(win_dist$ymax, n = -1))
win_dist$labelPosition <- (win_dist$ymax + win_dist$ymin) / 2

win_dist$label <- paste0(
  str_to_title(win_dist$winner), "\n",
  format(win_dist$count, big.mark = ",", scientific = FALSE), " (",
  round(win_dist$fraction * 100, 1), "%)"
)


ggplot(win_dist, aes(ymax = ymax, ymin = ymin, xmax = 4, xmin = 3, fill = winner)) +
  geom_rect(color = "black") +
  geom_text(x = 1.4, aes(y = labelPosition, label = label), size = 3.5) + 
  coord_polar(theta = "y") +
  xlim(c(-1, 4)) +
  scale_fill_manual(values = c("white" = "white", "black" = "black", "grey" = "grey")) +
  theme_void() +
  labs(
    title = "Games Won by Piece Color",
    fill = "Winner"
  ) +
    theme(legend.position = "none",
          plot.title = element_text(hjust = 0.5))
```

```{r}

chess |> 
  group_by(winner) |> 
  summarise(count = n()) |> 
  arrange(desc(count))

observed <- c(1001, 9107, 950)
expected <- c(9554, 9554, 950)
chisq.test(observed, p = expected / sum(expected))
```

The data suggests that **White has a significant advantage** when going first.

The chi-squared test with a p-value of 2.2e-16 indicates that White’s higher win rate (*10001 wins vs. 9107 for Black*) is unlikely to be due to random variation, and White likely has a real, statistically significant advantage by moving first.

------------------------------------------------------------------------

## Opening Moves

```{r}

opening_summary <- chess |> 
  filter(winner != "draw") |> 
  group_by(opening_name, winner) |> 
  summarise(count = n(), .groups = "drop") |> 
  arrange(desc(count))

top_openings <- opening_summary |> 
  group_by(opening_name) |> 
  summarise(total_count = sum(count)) |> 
  arrange(desc(total_count)) |> 
  slice_head(n = 10) |> 
  pull(opening_name)

top_opening_summary <- opening_summary |> 
  filter(opening_name %in% top_openings)

ggplot(top_opening_summary, aes(y = reorder(opening_name, count), x = count, fill = winner)) +
  geom_bar(stat = "identity", position = "dodge", color = "black") +
  scale_fill_manual(values = c("white" = "white", "black" = "black", "grey" = "grey")) +
  scale_y_discrete(expand = expansion(mult = c(0.05, 0.05))) +
  theme_minimal() +
  theme(
    legend.position = "none",
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    plot.title = element_text(hjust = 0.5)
  ) +
  labs(
    title = "Top Opening Moves by Winning color \n",
    x = "Count", 
    y = ""
  ) 
```
